The Design and Implementation of the Synchronous Language \CEU

### 2016-09-27 paper.tex
(>) notion of time

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} shows event occurrences over a continuous timeline.
The other boxes show how the same occurring events fit differently in each 
logical notion of time.
%
\begin{itemize}
\item \code{[Box-1]}:
\emph{Esterel with fixed-length ticks}~\cite{esterel.fixed}.
%
We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1} and invades 
\code{tick-2}, causing a \emph{timing violation}~\cite{esterel.fixed}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} can be considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
\emph{Esterel with variable-length ticks}~\cite{esterel.variable}.
This approach avoids the timing violation for \code{R(A)} and also results in
smaller idle periods because it adjusts the tick lengths to match the CPU times 
for the reactions.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} at \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered at \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
\emph{\CEU with unique and queue-based input events}.
We also assume a reaction \code{R(boot)} before any input.
%
Because the occurrence of event \code{A} is unique during \code{R(boot)}, the 
behavior in \CEU is similar to \code{Box-2} for its first two reactions 
(\code{tick-0} and \code{tick-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle periods in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
%
Note that, by definition, each reaction corresponds to an event occurrence of 
the same length.
\end{itemize}

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we consider that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
Esterel considers the events \code{B} and \code{C} to be simultaneous, even 
though \code{A} and \code{B} ``physically'' happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example to illustrate how to detect simultaneous 
button clicks in \CEU.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events in \CEU never overlap because they are atomic.
Automatic mutual exclusion simplifies reasoning about concurrency and is a 
prerequisite for the concurrency checks to be discussed in 
Section~\ref{sec.ceu.checks}.
\end{itemize}

>>>>

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} assumes an external observer with an absolute
reference clock that timestamps event occurrences over a
continuous timeline~\cite[Chapter~3]{es.time} (e.g., event \code{C} occurs at
$17ms521us$).
The other boxes show how the same occurring events fit differently in each 
logical notion of time.
%
\begin{itemize}
\item \code{[Box-1]}:
\emph{Esterel with fixed-length ticks}~\cite{esterel.fixed}.
%
Usually referred as \emph{sample-driven execution}~\cite{rp.twelve}.
%
We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1} and invades 
\code{tick-2}, causing a \emph{timing violation}~\cite{esterel.fixed}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} can be considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
\emph{Esterel with variable-length ticks}~\cite{esterel.variable}.
This approach avoids the timing violation for \code{R(A)} and also results in
smaller idle periods because it adjusts the tick lengths to match the CPU times 
for the reactions.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} at \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered at \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
\emph{\CEU with unique and queue-based input events}.
We also assume a reaction \code{R(boot)} before any input.
%
Because the occurrence of event \code{A} is unique during \code{R(boot)}, the 
behavior in \CEU is similar to \code{Box-2} for its first two reactions 
(\code{tick-0} and \code{tick-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle periods in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU reacts to the two instances of \code{E} independently, which are
handled in sequence.
%
%Note that, by definition, each reaction corresponds to an event occurrence of 
%the same length.
\end{itemize}

Sample-driven execution maps directly to hardware implementations and
involve no runtime queues.
%
However, considering the soft real-time application domain of \CEU, we
decided for the unique and queue-based semantics in \CEU for the reasons
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is implementation dependent:}
Tick lengths are not part of the program specification.
For instance, if we log the \code{Real World} box timeline and reproduce it in
implementations with different tick lengths, the behaviors might diverge.
%
\item \textbf{Events are never absolutely simultaneous:}
We consider that the notion of simultaneity should not be imposed by the
language, but defined explicitly for each use case (to be discussed in
Section~\ref{sec.ceu.timers}).
%In Section~\ref{sec.ceu.timers}, after introducing internal events and 
%first-class timers, we show an example to illustrate how to detect simultaneous 
%button clicks in \CEU.
In tick-based approaches, simultaneity depends on the length of discrete ticks.
For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
events \code{B} and \code{C} are simultaneous, even though \code{A} and
\code{B} ``physically'' happen much closer to one another.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to exclusive events are atomic and never overlap.
Automatic mutual exclusion simplifies reasoning about concurrency and is a 
prerequisite for the concurrency checks to be discussed in 
Section~\ref{sec.ceu.checks}.
\end{itemize}

Note that Esterel supports declarations for mutually exclusive inputs which
cannot be simultaneous in the same reaction
(e.g., \code{relation A\#B\#C\#D\#E;})~\cite{esterel.primer}.
Making all inputs mutually exclusive and adopting the variable-length-tick
approach is equivalent to \CEU's notion of time, which can be seen as an
imposed restriction on the semantics of Esterel.
